### 2013-05-16 context-dependence.tex
.

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}

@@

%%%%
\subsection{A likelihood function}

Now we know how to compute the (marginal) probability that a particular subsequence changes into another subsequence.
For patterns $x$ and $y$, with offset $\ell$ (so that $|x|=|y|+2\ell$),
we can count how many times $x$ appears in $X(0)$,
and how many times $x$ appears in $X(0)$ while $y$ appears at the same place in $X(t)$, offset by $\ell$.
For instance, if $x$ is ACTCAC, $y$ is CGC, and $\ell=3$, this looks like:
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$  
\end{tabular}
\end{center}
where $\cdot$ could be any base.
Let $N(x)$ and $N_\ell(x,y)$ count these events, respectively.

From the definitions above, 
the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x)p_{\ell,m,r}(x,y)$.
Covariances are also easy to compute.
If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{|x|}(0)=x$ and $X_i^({|y|})(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then
\begin{align}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\ell,m}(x,y)(1-p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\ell,m}(x,y)^2 \right) \\
  &= n p_{\ell,m}(x,y)(1-(1+2m+4\ell)p_{\ell,m}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)] 
    & \approx - n (2m+4\ell) p_{\ell,m}(x_1,y_1)p_{\ell,m}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align}
